Recursive Solutions to First-Order Model Counting

### 2023-01-31 doc/paper/paper.tex
1/4 proofs complete

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usetikzlibrary{cd}
\usetikzlibrary{shapes}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{proposition}{Proposition}
\newtheorem{claim}{Claim}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. Lifted inference
algorithms are typically used on probabilistic models such as probabilistic
programming
languages~\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17},
Markov logic
networks~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical~\cite{DBLP:journals/ml/KimmigMG15} and statistical
relational~\cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \FOtwo{} with tree
axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On
the empirical front, there are several open-source implementations of exact
WFOMC algorithms: \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving~\cite{DBLP:journals/cacm/GogateD16}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}.
% TODO: add FastWFOMC (only for \FOtwo{})

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy\footnote{The algorithm is not described in any paper but can be
  found in its source code at \url{https://dtai.cs.kuleuven.be/drupal/wfomc}}
search, \textsc{Crane} uses a hybrid search algorithm that applies some rules
greedily and some using breadth-first search. Second, the product of compilation
is not directly evaluated but rather interpreted as a collection of functions on
domain sizes.

@@

\begin{proposition}[Correctness of $\GDR$]
  Let $\phi$ be the formula used as input to \cref{alg:domainrecursion} and
  $\phi'$ be one of the formulas constructed by the algorithm. Then
  $\phi \equiv \phi'$.
\end{proposition}
\begin{proof} % TODO: finish (Proof technique: if one is true then the other one will be true and vice versa)
  Let $\Omega \in \mathcal{D}$ be the domain selected by
  \cref{line:condition} of the algorithm.
  \begin{claim}\label{claim:logic}
  \[
    \forall X \in \Omega\text{. } \psi \equiv \psi[x/\{\, X \,\}] \land \forall X \in \Omega\text{. } X \ne x \implies \psi
  \]
  \end{claim}
  \begin{claim}
    Iteration of \cref{claim:logic} for all variables in some set of variables
    $V$ produces the same clauses as subset enumeration on
    \cref{line:conditions}.
  \end{claim}
\end{proof}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\CR$ nodes.}\label{alg:constraintremoval}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{replaceable pair $(\Omega \in \mathcal{D}, x \in \Omega)$\label{line:crconditions}}{
    add a new domain $\Omega'$ to $\mathcal{D}$\;
    $\phi' \gets \emptyset$\;
    \ForEach{clause $(L, C, \delta) \in \phi$}{
      $C' \gets \{\, (a, b) \in C \mid b \ne x \,\}$\;\label{line:constraintremoval}
      \nosemic$\delta' \gets X \mapsto
      \begin{cases}
        \Omega' & \text{if } \delta(X) = \Omega\\
        \delta(X) & \text{otherwise;}
      \end{cases}$\;\label{line:newdelta}
      $\phi' \gets \phi' \cup \{\, (L, C', \delta') \,\}$\;
    }
    $S \gets S \cup \{\, (\CR_{\Omega \mapsto \Omega'}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{proposition}[Correctness of $\CR$]
\end{proposition}
\begin{proof} % TODO: finish (Proof technique: if one is true then the other one will be true and vice versa)
  \[
    \forall X \in \Omega \ni x\text{. } X \ne x \implies \psi \equiv \forall X \in \Omega \setminus \{\, x \,\}\text{. } \psi
  \]
\end{proof}

@@

% TODO: maybe something other than MC? Introduce the notation and the role of \sigma.
% TODO: add an argument about why this recursion is (or can easily be made) finite
\begin{proposition}[Correctness of $\Reff$]
\end{proposition}
\begin{proof} % TODO: finish (Proof technique: show a bijection between the sets of models)
  For a given formula $\phi$, \cref{alg:trycache} attempts to find a formula
  $\psi$ and a map $\rho\colon \Doms(\psi) \to \Doms(\phi)$ such that
  $\MC(\psi, \sigma \circ \rho\colon \Doms(\psi) \to \mathbb{N}_{0}) = \MC(\phi, \sigma\colon \Doms(\phi) \to \mathbb{N}_{0})$.
\end{proof}

@@

\section{Empirical Results}\label{sec:results}
% TODO: when describing experimental results, say that a solution for the last
% (unsolved) instance can be found, but is too complicated to evaluate by hand.
% Add that the linear order axiom is also liftable. Cite the same paper in the
% introduction as well.

% TODO: make sure this table floats to the right page and look at other floating
% objects too (e.g., is [t] on?)
\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count
various kinds of functions. This class of instances is chosen because of its
simplicity and ubiquity and the inability of state-of-the-art WFOMC algorithms
to solve many such instances. Note that other WFOMC
algorithms---\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} and probabilistic
theorem proving~\cite{DBLP:journals/cacm/GogateD16}---are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

@@

% the process: both for running algorithms and for determining complexities
In our experiments, we consider all sixteen combinations of these properties,
i.e., injectivity, surjectivity, partiality, and endo-. We run each algorithm
once on each of the sixteen instances.\footnote{While domain sizes have to be
  specified as part of the input, they are immaterial to the search procedure.}
\textsc{ForcLift} is always run until it terminates; \textsc{Crane} is run until
either five solutions are found or the search tree reaches height
six.\footnote{The search tree has a high branching factor, so exploring all
  nodes at depth five takes at most a few seconds, whereas doing the same for
  depth six can be computationally infeasible in some cases.} If successful,
\textsc{ForcLift} generates a circuit, and \textsc{Crane} generates one or more
(complete) FCGs. In both cases, we manually convert the resulting graphs into
definitions of functions as described in \cref{sec:interpret}. We then assess
the complexity of each solution by hand and pick the best in case \textsc{Crane}
returns several solutions of varying complexities.

@@

These results are summarised in \cref{tbl:results}, where we compare the
solutions found by both algorithms to the best known ways of computing the same
quantities. The best-known asymptotic complexity for computing total surjections
is by Earnest~\shortcite{30049}. All other best-known complexity results are
inferred from the formulas and programs on the on-line encyclopedia of integer
sequences~\cite{oeis}. On instances that could already be solved by
\textsc{ForcLift}, the two algorithms perform equally well. However,
\textsc{Crane} can also solve all but one instances that \textsc{ForcLift} fails
on in at most cubic time. See \cref{app:solutions} for the exact solutions
produced by \textsc{Crane} that correspond to the complexity results in
\cref{tbl:results}.

@@

% A summary of what's been done/achieved and how

@@

% TODO: cite Lifted Reasoning for Combinatorial Counting

@@

% TODO: fix the introductory description

whose complexity is in \cref{tbl:results} (including those findable by
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}). Both before and after
algebraic simplification, excluding multiplication by one. All of these
solutions have been checked to be correct when combined with suitable base
cases.

>>>>

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usetikzlibrary{cd}
\usetikzlibrary{shapes}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
% \newtheorem{theorem}{Theorem}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. Lifted inference
algorithms are typically used on probabilistic models such as probabilistic
programming
languages~\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17},
Markov logic
networks~\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical~\cite{DBLP:journals/ml/KimmigMG15} and statistical
relational~\cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC has received more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. There is also a recent line
of work that uses FOMC-based methods to count solutions to constraint
satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23}.

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \FOtwo{} with tree
axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On
the empirical front, there are several open-source implementations of exact
WFOMC algorithms: \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11},
probabilistic theorem proving~\cite{DBLP:journals/cacm/GogateD16},
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}, and
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}. Approximate counting is
supported by \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy\footnote{The algorithm is not described in any paper but can be
  found in its source code at \url{https://dtai.cs.kuleuven.be/drupal/wfomc}}
search, \textsc{Crane} also supports a hybrid search algorithm that applies some
rules greedily and some using breadth-first search. Second, the product of
compilation is not directly evaluated but rather interpreted as a collection of
functions on domain sizes.

@@

\begin{theorem}[Correctness of $\GDR$]
  Let $\phi$ be the formula used as input to \cref{alg:domainrecursion} and
  $\phi'$ be one of the formulas constructed by the algorithm. Then
  $\phi \equiv \phi'$.
\end{theorem}
\begin{proof} % TODO: finish (Proof technique: if one is true then the other one will be true and vice versa)
  Let $\Omega \in \mathcal{D}$ be the domain selected by
  \cref{line:condition} of the algorithm.
  \begin{claim}\label{claim:logic}
  \[
    \forall X \in \Omega\text{. } \psi \equiv \psi[x/\{\, X \,\}] \land \forall X \in \Omega\text{. } X \ne x \implies \psi
  \]
  \end{claim}
  \begin{claim}
    Iteration of \cref{claim:logic} for all variables in some set of variables
    $V$ produces the same clauses as subset enumeration on
    \cref{line:conditions}.
  \end{claim}
\end{proof}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\CR$ nodes.}\label{alg:constraintremoval}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{replaceable pair $(\Omega \in \mathcal{D}, x \in \Omega)$\label{line:crconditions}}{
    add a new domain $\Omega'$ to $\mathcal{D}$\;\label{line:newdomain}
    $\phi' \gets \emptyset$\;
    \ForEach{clause $(L, C, \delta) \in \phi$}{
      $C' \gets \{\, (a, b) \in C \mid b \ne x \,\}$\;\label{line:constraintremoval}
      \nosemic$\delta' \gets X \mapsto
      \begin{cases}
        \Omega' & \text{if } \delta(X) = \Omega\\
        \delta(X) & \text{otherwise;}
      \end{cases}$\;\label{line:newdelta}
      $\phi' \gets \phi' \cup \{\, (L, C', \delta') \,\}$\;
    }
    $S \gets S \cup \{\, (\CR_{\Omega \mapsto \Omega'}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{theorem}[Correctness of $\CR$]
  Let $\phi$ be the input formula of \cref{alg:constraintremoval}, $(\Omega, x)$
  a replaceable pair, and $\phi'$ the formula constructed by the algorithm, when
  $(\Omega, x)$ is selected on \cref{line:crconditions}. Then
  $\phi \equiv \phi'$, where the domain $\Omega'$ introduced on
  \cref{line:newdomain} is interpreted as $\Omega \setminus \{\, x \,\}$.
\end{theorem}
\begin{proof}
  Since there is a natural bijection between the clauses of $\phi$ and $\phi'$,
  we shall argue about the equivalence of each pair of clauses. Let $c$ be an
  arbitrary clause of $\phi$ and $c'$ its corresponding clause of $\phi'$.

  If $c$ has no variables with domain $\Omega$, then it cannot have any
  constraints involving $x$, so $c' = c$. Otherwise, for notational simplicity,
  let us assume that $X$ is the only variable in $c$ with domain $\Omega$ (the
  proof for an arbitrary number of variables is virtually the same). By
  \cref{def:replaceable}, we can rewrite $c$ as $\forall X \in \Omega\text{.
  } X \ne x \implies \psi$, where $\psi$ is a formula with $X$ as the only free
  variable and with no mention of either $x$ or $\Omega$. Then
  $c' \equiv \forall X \in \Omega'\text{. } \psi$. Since
  $\Omega' \coloneqq \Omega \setminus \{\, x \,\}$, we have that $c \equiv c'$.
  Since $c$ was an arbitrary clause of $\phi$, this completes the proof that
  $\phi \equiv \phi'$.
\end{proof}

@@

% TODO: maybe something other than MC? Introduce the notation and the role of \sigma.
% TODO: add an argument about why this recursion is (or can easily be made) finite
\begin{theorem}[Correctness of $\Reff$]
\end{theorem}
\begin{proof} % TODO: finish (Proof technique: show a bijection between the sets of models)
  For a given formula $\phi$, \cref{alg:trycache} attempts to find a formula
  $\psi$ and a map $\rho\colon \Doms(\psi) \to \Doms(\phi)$ such that
  $\MC(\psi, \sigma \circ \rho\colon \Doms(\psi) \to \mathbb{N}_{0}) = \MC(\phi, \sigma\colon \Doms(\phi) \to \mathbb{N}_{0})$.
\end{proof}

@@

\begin{theorem}
  Any collection of functions returned by \textsc{Crane} can be evaluated in
  time polynomial in domain sizes.
\end{theorem}
\begin{proof}% TODO (I have some handwritten notes)
  % prove that all functions output by Crane can be computed in polytime (must, should be easy: 1) bound the number of possible arguments that any function can have, 2) prove that the computation time of any dynamic programming cell is always polytime (what about functions depending on other functions?))
\end{proof}

\section{Empirical Results}\label{sec:results}
% TODO: Add that the linear order axiom is also liftable. Cite the same paper in the
% introduction as well.

% TODO: make sure this table (as well as other floating objects) is on the right page
\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{4}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \cmark & \xmark & & \multicolumn{3}{c}{Same as injections from $\Delta$ to $\Gamma$} \\
    \cmark & \cmark & & \multicolumn{3}{c}{Same as endo-injections} \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & \multirow{3}{*}{Bijections} & $m$ & --- & $m$ \\
    \xmark & \cmark & & \multicolumn{3}{c}{\multirow{2}{*}{Same as (partial) (endo-)injections}} \\
    \cmark & \cmark/\xmark & & \multicolumn{3}{c}{} \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: It is a standard to discuss a few lifted theories like the smokers and
% friends example. Can you also include some empirical results on such theories
% as well to appeal to the sanity of the reader that it actually works for
% problems in frameworks such as Markov log networks

We compare \textsc{Crane} and
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} on their ability to count
various kinds of functions. This class of instances is chosen because of its
simplicity and ubiquity and the inability of state-of-the-art WFOMC algorithms
to solve many such instances. Note that other WFOMC
algorithms---\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}, probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}, and
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}---are unable to solve any of
the instances that \textsc{ForcLift} fails on. We begin by describing how such
function-counting problems can be expressed in FOL\@. \textsc{ForcLift} then
translates these sentences in FOL to formulas as defined in \cref{def:formula}.

@@

% the process: both for running algorithms and for determining complexities
In our experiments, we consider all sixteen combinations of these properties,
i.e., injectivity, surjectivity, partiality, and endo-. We run each algorithm
once on each of the sixteen instances.\footnote{While domain sizes have to be
  specified as part of the input, they are immaterial to the search procedure.}
\textsc{Crane} is run in hybrid search mode until either five solutions are
found or the search tree reaches height six.\footnote{The search tree has a high
  branching factor, so exploring all nodes at depth five takes at most a few
  seconds, whereas doing the same for depth six can be computationally
  infeasible in some cases.} \textsc{ForcLift} is always run until it
terminates. If successful, \textsc{ForcLift} generates a circuit, and
\textsc{Crane} generates one or more (complete) FCGs. In both cases, we manually
convert the resulting graphs into definitions of functions as described in
\cref{sec:interpret}. We then assess the complexity of each solution by hand and
pick the best in case \textsc{Crane} returns several solutions of varying
complexities.

@@

These results are summarised in \cref{tbl:results}, where we compare the
solutions found by both algorithms to the best known ways of computing the same
quantities. The best-known asymptotic complexity for computing total surjections
is by Earnest~\shortcite{30049}. All other best-known complexity results are
inferred from the formulas and programs on the on-line encyclopedia of integer
sequences~\cite{oeis}. On instances that could already be solved by
\textsc{ForcLift}, the two algorithms perform equally well. However,
\textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See \cref{app:solutions} for
the exact solutions produced by \textsc{Crane} that correspond to the complexity
results in \cref{tbl:results}.

@@

@@

@@

In this appendix, we list the exact function definitions produced by
\textsc{Crane} for all of the problem instances in \cref{sec:results} both
before and after algebraic simplification (excluding multiplications by one).
The correctness of all of them has been checked by identifying suitable base
cases and verifying the numerical answers across a range of domain sizes.

\begin{enumerate}
  \item $\Theta(m)$ solution for counting $\Gamma \to \Gamma$ functions:
  \[
    f(m) = {\left(-1 + \sum_{l=0}^{m} \binom{m}{l} [l < 2]\right)}^{m} = m^{m}.
  \]
  \item $\Theta(m^3 + n^3)$ solution for counting $\Gamma \to \Delta$
  surjections:
  \begin{align*}
    f(m, n) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k}\\
               &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
    ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{n} \binom{n}{k} {(-1)}^{n-k} {(k+1)}^{l}.
  \end{align*}
  \item $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ surjections:
  \begin{align*}
    f(m) ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k}\\
            &{\left( \sum_{j=0}^{k} \binom{k}{j} [j < 2] \right)}^{l} \\
    ={}& \sum_{l=0}^{m} \binom{m}{l}{(-1)}^{m-l} \sum_{k=0}^{m} \binom{m}{k} {(-1)}^{m-k} {(k+1)}^{l}.
  \end{align*}
  \item $\Theta(mn)$ solution for counting $\Gamma \to \Delta$ injections and
  partial injections (with different base cases):
  \begin{align*}
    f(m, n) ={}& \sum_{l=0}^m \binom{m}{l} [l<2] f(m-l, n-1)\\
    ={}& f(m, n-1) + mf(m-1, n-1).
  \end{align*}
  \item $\Theta(m^{3})$ solution for counting $\Gamma \to \Gamma$ injections:
  \begin{align*}
    f(m) ={}& \sum_{l=0}^{m} \binom{m}{l} {(-1)}^{m-l} g(m, l); \\
    g(m, l) ={}& \sum_{k=0}^{l} \binom{l}{k} [k < 2] g(m - 1, l - k)\\
    ={}& g(m - 1, l) + lg(m - 1, l - 1).
  \end{align*}
  % \item $\Theta(m^{2})$ solution for counting $\Gamma \to \Gamma$ partial
  %       injections:
  % \begin{align*}
  %   f(m) ={}& f(m - 1) + \sum_{l = 0}^{m} \binom{m}{l} \sum_{k = 0}^{m} \binom{m}{k}\\
  %           &[1 \le l < 3] [k < 2] 2^{m - l} g(m - l, m - k - 1)\\
  %   ={}& f(m - 1) + m2^{m-1}g(m-1, m-1)\\
  %   +{}& m^{2}2^{m-1}g(m-1, m-2)\\
  %   +{}& \binom{m}{2}2^{m-2}g(m-2, m-1) \\
  %   +{}& \binom{m}{2}m2^{m-2}g(m-2, m-2)\\
  %   g(i, j) ={}& \sum_{h = 0}^{j} \binom{j}{h} [h < 2] g(i - 1, j - h)\\
  %   ={}& g(i-1, j) + jg(i-1, j-1).
  %   \end{align*}
  \item $\Theta(m)$ solution for counting $\Gamma \to \Delta$ bijections:
  \[
    f(m, n) = mf(m-1, n-1).
  \]
\end{enumerate}
